---
title: "Monitoring Phytophthora cinnamomi disease incidence"
output:
  html_document:
    df_print: paged
---


```{r message=FALSE, warning=FALSE}
library(ggmap)
library(tidyverse)
library(ggplot2)
library(readr)
library(glmmTMB)
library(DHARMa)
library(pROC)
library(emmeans)
library(lme4)
library(car)
library(wesanderson)
set.seed(22052024)
```


## Data

- we combine years 2006-2008, and hereafter these are grouped as 2008
- we combine landforms as follows 

 i. ridgetops
 ii. upper and mid-slopes, 
 iii. lower-slopes and in-filled valleys and 
 iv. rises or long gentle foot slopes and alluvial deposits

For analysis stability, we scale distance to road and watercourse. We unscaled variables later for interpretation. 

We remove one site with unknown vegetation type. 

Please copy data from XX into data folder. 


```{r}
datadir <- "../data"

df_phyto <- read.csv(file.path(datadir,"Sthern_Syd_Pc_compl_updated5.csv"), strip.white = TRUE,
                     na.strings = c("", "NA")) %>% 
  mutate(Test = ifelse(PC_present %in% c( "Positive", "positive"), 1, 0),
         Year = ifelse(Year%in% c(2006, 2007), 2008, Year), 
         Year = factor(Year)) 


df_phyto_clean <- df_phyto %>% 
  transmute(Code = factor(Code), 
         Year,
         Test,
         Scivi_combined_updated = factor(Scivi_combined_updated),
         Landform_ordinal_cat = case_when(Landform_ordinal %in% c(1)~ "1",
                                          Landform_ordinal %in% c(2,3)~ "2-3",
                                          Landform_ordinal %in% c(4,5)~ "4-5",
                                          Landform_ordinal %in% c(6,7)~ "6-7"),
         Landform_ordinal = factor(Landform_ordinal), 
         s_DISTROAD_cs = (DISTROAD_cs - 200)/300,
         s_DISTHYDRO_cs = (DISTHYDRO_cs  - 100)/200,
         Longitude, Latitude,
         Location = substr(Collection_area,1,8)) %>% 
  filter(!is.na(Scivi_combined_updated))

#for model with spatial random effect
df_phyto_clean$pos <- numFactor(df_phyto_clean$Longitude, df_phyto_clean$Latitude)
```

Correlation matrix
```{r}
df_phyto_clean %>% 
  select(s_DISTROAD_cs, s_DISTHYDRO_cs) %>% 
  cor()
```

## Models

```{r}
phyto_mod_nonspatial <- glmmTMB(Test ~ s_DISTROAD_cs + s_DISTHYDRO_cs +
                                  Location + Year  + Landform_ordinal_cat +
                                  (1|Scivi_combined_updated) + (1 | Code) ,
                                  data = df_phyto_clean,
                                family = binomial,
                                control = glmmTMBControl(rank_check = "adjust"))

phyto_mod_spatial <- update(phyto_mod_nonspatial, .~. + exp(pos + 0 | Year))

```

We do not include an interaction for year and location as only one location (Dharawal Reserves) was sampled in multiple years. 


```{r}
AIC(phyto_mod_spatial, phyto_mod_nonspatial)
```

The spatial model has better (smaller) AIC, so this is our primary model.


## Inference

### Model with the spatial random effect

```{r}
drop1(phyto_mod_spatial, test="Chisq") # type 2 anova
```


```{r}
emmeans(phyto_mod_spatial, ~ Year | Location, type = "response")
```

### Sensitivity analysis, removing the spatially structured random effect

```{r}
drop1(phyto_mod_nonspatial, test="Chisq")# type 2 anova
```

Rescale odds ratio for distance to watercourse to per 100m for easier interpretation

```{r}
1/exp(confint(phyto_mod_nonspatial, parm = c("s_DISTHYDRO_cs"))/200*100)
```
For each 100 metres closer to a watercourse, the odds of P.cinnamomi increase by 19% (95% CI: 2%-39%, p = 0.010)


### Plot


```{r}
pos = position_dodge(width=0.3)

em_spatial <- emmip(phyto_mod_spatial, ~ Year | Location, CIs = TRUE, type = "response", 
                       plotit = FALSE) %>% 
  mutate(model = "spatial")

plot_phyt3 <- em_spatial %>% 
  filter(!(Year!=2008  & Location != "Dharawal")) %>%
  ggplot(aes(Year, yvar, color = Location)) +
  geom_point(position = pos, size = 2) + 
  geom_path(position = pos , size = 1) +
  geom_errorbar(aes(ymin = LCL, ymax = UCL),
                position = pos,
                alpha = 0.5, width = 0, size = 2) + 
  guides(fill=guide_legend(title=NULL))+
  ylim(0,1)+
 theme(axis.text = element_text(size = 12))+
  theme(axis.title = element_text(size = 14))+
  theme(plot.title = element_text(size = 14))+
  theme(strip.text = element_text(size = 14))+
  theme(legend.text = element_text(size = 12))+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black")) +
  #theme(legend.position = "none") +
  theme(legend.position = "bottom", legend.key = element_rect(fill = NA))+
  
  theme(plot.margin = unit(c(0.2,2,0.2,0.2), "cm"))+
  theme(legend.title = element_blank())+
    ylab(expression(atop(paste("Probability of ",italic("Phytophthora cinnamomi")), paste(" infection (+/-95% CI)"))))+
  scale_fill_discrete(labels=c('Dharawal parks', 'Georges River/ Botany Bay parks', 'Royal / Heathcote / Garawarra parks'))+
  #scale_color_manual(values  = wes_palette(n=3, name="Moonrise2"))
  scale_color_manual(labels = c('Dharawal', 'Georges River/ Kamay Botany Bay', 'Royal / Heathcote / Garawarra'), values  = c("darkgoldenrod4", "darkgoldenrod2", "gold"))
plot_phyt3
#ggsave(plot = plot_phyt3, file = "../plots/phyt_plotTM9_updated.tiff" , width = 180, height = 150, units = "mm", device = "tiff")
```



