---
title: "Inconsistent pathogen detection and the role of sentinel species"
output:
  html_document:
    df_print: paged
---



```{r message=FALSE, warning=FALSE}
library(ggmap)
library(tidyverse)
library(ggplot2)
library(readr)
library(glmmTMB)
library(DHARMa)
library(pROC)
library(emmeans)
library(lme4)
```



## Data

Please copy Sthern_Syd_Pc_compl_updated5.csv from https://doi.org/10.5281/zenodo.14399365 into data folder. 


```{r}
datadir <- "../data"

df_phyto <- read.csv(file.path(datadir,"Sthern_Syd_Pc_compl_updated5.csv"), strip.white = TRUE) %>% 
   filter(Site_severity != "NA") %>% 
  #filter(Year!="2022") %>%
  mutate(Test = ifelse(Phyto_genus_present %in% c( "Positive", "positive"), 1, 0),
         Year = ifelse(Year == 2007, 2008, Year),
         Year = factor(Year)) 

```

## Visual assessment - full dataset

This assumes a cut-off of 1 (i.e. 0 vs 1,2 or 3) for Site_severity

```{r}
df_phyto %>% 
  mutate(Field = ifelse(Site_severity == 0, 0, 1)) %>% # make site severity binary
  select(Test, Field) %>%
  table() %>% 
  prop.table(margin = 1)
```

Visual field assessment and laboratory assays have inconsistent detections when using the full dataset. 

```{r}
pROC_obj <- roc(df_phyto$Test,
                df_phyto$Site_severity,
            # arguments for ci
            ci=TRUE,
            # arguments for plot
            plot=TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, grid=TRUE,
            print.auc=TRUE)
```


##Visual assessment - Xanthorrhoea as sentinel species

Only include data where Xanthorrhoea was present and run the analysis again (cut down dataset below).



```{r}
datadir <- "../data"



df_phyto_xanth <- read.csv(file.path(datadir,"Sthern_Syd_Pc_compl_updated5_xanthorr.csv"), strip.white = TRUE) %>%
filter(Site_severity != "NA") %>% 
  #filter(Year!="2022") %>%
  mutate(Test = ifelse(Phyto_genus_present %in% c( "Positive", "positive"), 1, 0),
         Year = ifelse(Year == 2007, 2008, Year),
         Year = factor(Year))
```


This assumes a cut-off of 1 (i.e. 0 vs 1,2 or 3) for Site_severity

```{r}
df_phyto_xanth %>% 
  mutate(Field = ifelse(Site_severity == 0, 0, 1)) %>% # make site severity binary
  select(Test, Field) %>%
  table() %>% 
  prop.table(margin = 1)
```


Visual field assessment is more consistent with laboratory assay when Xanthorrhoea is a community component. 


```{r}
pROC_obj <- roc(df_phyto_xanth$Test,
                df_phyto_xanth$Site_severity,
            # arguments for ci
            ci=TRUE,
            # arguments for plot
            plot=TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, grid=TRUE,
            print.auc=TRUE)
```





